### 2018-07-05 paper/4-build.tex
More from Simon

\vspace{1mm}
We have bolded two different phrases above and, as we will see, the part of the
build system responsible for scheduling tasks in the dependency order
(a `scheduler') can be cleanly separated from the part responsible for deciding
whether a key needs to be rebuilt (a `rebuilder'). We therefore tackle each
aspect separately in subsections~\S\ref{sec-dependency-orderings}
and~\S\ref{sec-out-of-date}.

\subsection{Respecting the dependency order}
\label{sec-dependency-orderings}

@@

To handle dynamic dependencies we can use the following speculative approach:
build tasks in an arbitrary order, discovering their dependencies on the fly;
whenever a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task,
and switch to building the dependency \hs{dep}; eventually the previously
aborted task is restarted and makes further progress thanks to \hs{dep} now
being up to date. When a build completes, we can record the resulting order,
using it as a starting point for the next build and fixing it up if a task's
dependencies have changed, as \Excel does.

This approach requires a way to abort tasks that have failed due to out-of-date
dependencies. It is also not minimal in the sense that a task may start, do some
meaningful work, and then abort.

\Bazel's restarting scheduler does not store the discovered order between build
runs; instead, it stores the most recent task dependency information. Since this
information may become outdated, \Bazel may also need to abort a task if a
newly discovered dependency is out-of-date.

@@

\subsection{Determining out-of-date keys} \label{sec-out-of-date}

@@

The idea of a dirty bit is to have one piece of persistent information per key,
saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
are set to clean. When the next build starts, anything that changed between the
two builds is marked dirty. If a key and all its transitive dependencies are
clean, the key does not need to be rebuild.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it, and
also making sure all cells that depend on it are recursively marked dirty too.
To mark dependent cells dirty, it is necessary to statically know task
dependencies. This is possible only for applicative tasks (see \S\ref{sec-deps}).
For monadic tasks, \Excel computes a \emph{static approximation} of task
dependencies: for functions such as \cmd{IF}, the approximation includes all
cells that are mentioned in the formula (the \cmd{IF}'s condition and both
branches), whereas functions such as \cmd{INDIRECT} whose dependencies cannot be
guessed, are conservatively assumed to depend on all cells and are therefore
always marked dirty.

>>>>

\vspace{1mm}
We have emboldened two different aspects above: the part of the
build system responsible for scheduling tasks in the dependency order
(a `scheduler') can be cleanly separated from the part responsible for deciding
whether a key needs to be rebuilt (a `rebuilder'). We tackle each
aspect separately in subsections~\S\ref{sec-dependency-orderings}
and~\S\ref{sec-out-of-date}.

\subsection{The scheduler: respecting the dependency order}
\label{sec-dependency-orderings}

@@

To handle dynamic dependencies we can use the following approach:
build tasks in an arbitrary initial order, discovering their dependencies on the fly;
whenever a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task,
and switch to building the dependency \hs{dep}; eventually the previously
aborted task is restarted and makes further progress thanks to \hs{dep} now
being up to date.
This approach requires a way to abort tasks that have failed due to out-of-date
dependencies. It is also not minimal in the sense that a task may start, do some
meaningful work, and then abort.

To reduce the number of aborts (often to zero) Excel records the
discovered task order in its \emph{calc chain}, and uses it as the
starting point for the next build (\S\ref{sec-background-excel}).
\Bazel's restarting scheduler does not store the discovered order
between build runs; instead, it stores the most recent task dependency
information. Since this information may become outdated, \Bazel may
also need to abort a task if a newly discovered dependency is
out-of-date.

@@

\subsection{The rebuilder: determining out-of-date keys} \label{sec-out-of-date}

@@

The idea of a dirty bit is to have one piece of persistent information per key,
saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
are set to clean. When the next build starts, anything that changed between the
two builds is marked dirty. If a key and all its transitive dependencies are
clean, the key does not need to be rebuilt.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it.
It also marks dirty all cells that depend (transitively) on the modified cell.
Excel does not record the dynamic dependencies of each cell,
so instead it computes a \emph{static over-approximation} -- it is always
safe for it to make more cells dirty than is strictly necessary, but not
vice versa.  The over-approximation is as follows: a cell is marked dirty
if its formula statically refers to a dirty cell, or if
the formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
guessed from the formula alone.  The over-approximation is clear for \cmd{INDIRECT},
but it is also present for \cmd{IF}, where both branches are followed even though
dynamically only one is used.
